Destruction of metrizability in generalized inverse limits

### 2015-09-22 articles/destroying-metrizability-in-generalized-inverse-limits/article.tex
scott's revisions

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and intervals on a linearly ordered space, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$ The \emph{identity on} $X$ is the u.s.c. function $\iota$ given by $\iota(x) = \{x\}$ for all $x \in X$; we may simply say the \emph{identity} (or, more simply, $\iota$) when the space $X$ is understood from context.

@@

Because it can be difficult to check if a large collection of u.s.c. functions is exact, a much more tractable case would be that of an inverse limit system with a single idempotent bonding function $f$. That is, suppose $X$ is a non-empty compact Hausdorff space and $f: X \rightarrow 2^{X}$ is an idempotent u.s.c. function. If $X_{\alpha} = X$ for each $\alpha \in \kappa$ and $f_{\alpha \beta} = f$ for each $\alpha < \beta \in \kappa$, then $\{X_{\alpha}, f_{\alpha \beta}, \kappa\} = \{X, f, \kappa\}$ is an \emph{inverse limit system with the single bonding function $f$}, and we denote the inverse limit of the system as $\varprojlim \{X, f, \kappa\}$. (Of course, since $f$ is idempotent, the collection $\textbf{f}$ of bonding functions is automatically exact.) We note that, although the only continuous idempotent surjective function from a compact Hausdorff space $X$ to itself is the identity map, there are many non-trivial examples of u.s.c. idempotent surjective functions on $X$; see Figure \ref{idempotentUsc} for examples.

@@

Ingram and Mahavier showed in \cite{i m paper} that $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \omega\}$ is non-empty and compact. They also found that
if $f_{\alpha \beta}$ is always continuum-valued, then $\varprojlim \{X_{\alpha}, f_{\alpha \beta}, \omega\}$ is also connected, and therefore a Hausdorff continuum. Varagona generalized these results for other index sets (including ordinals $\kappa$) in Theorems 2.1 and 2.2 of \cite{varagona}. Greenwood and Kennedy showed in \cite{greenwood kennedy} that there is a non-continuum-valued u.s.c. surjective idempotent bonding map $f$ for which $\varprojlim \{X_{\alpha}, f, \omega\}$ is disconnected.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{idempotent_usc_4.pdf}
\end{center}
\caption{Non-trivial u.s.c. surjections satisfying condition $\Gamma$}
\label{conditionGamma}
\end{figure}

@@

Note that while we have assumed that $f:[0,1]\to C([0,1])$ (which also guarantees that the inverse limit is a continuum), it is an open question whether $\Gamma$ must also hold for any surjective idempotent u.s.c. $f:[0,1]\to 2^{[0,1]}$ (regardless if the inverse limit remains connected).

@@

Note that when $\kappa\geq\omega_1$, $\omega_1\in\kappa+1$ is a point of non-first-countability; therefore any space containing $\kappa+1$ as a subspace is nonmetrizable.

@@

All compact metric spaces are embeddable in $\mathbb R^\omega=\Sigma\mathbb R^\omega$ and therefore Corson compact. In addition, every closed subspace of a Corson compact is Corson compact. While $\Sigma\mathbb R^\kappa$ is not first-countable for $\kappa\geq\omega_1$, any subspace holds the weaker notion of being a $W$-space as defined by Gruenhage in \cite{gruenhage}. However, $\omega_1+1$ does not, as every countable sequence of ordinals less than $\omega_1$ is bounded above and therefore cannot converge to $\omega_1$ itself; therefore any space containing $\kappa+1$ for $\kappa\geq\omega_1$ cannot even be Corson compact.

@@

\bibitem{greenwood kennedy} Greenwood, Sina and Kennedy, Judy, {\it Connected generalized inverse limits}, Topology and its Applications,
159 (2012), no. 1, 57–68.

\bibitem{gruenhage} G. Gruenhage, {\it Infinite games and generalizations of first-countable spaces}, Gen. Topol. Appl., 6 (1976), pp. 339–352

>>>>

Suppose $X$ and $Y$ are non-empty compact Hausdorff spaces. Then the set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{upper semi-continuous} (u.s.c.) at $x \in X$ if, for every open set $V$ in $Y$ with $f(x) \subseteq V$, there exists an open set $U$ in $X$ with $x \in U$ and $f(u) \subseteq V$ for each $u \in U$. If $f$ is u.s.c. at each $x \in X$, we simply say \emph{$f$ is u.s.c.} If $f: X \rightarrow 2^{Y}$ is a set-valued function, then the \emph{graph of $f$}, denoted $G(f)$, is given by $G(f) = \{\langle x , y \rangle \in X \times Y \ | \ y \in f(x)\}$. (Note that, to help distinguish between ordered pairs and intervals on a linearly ordered space, we will use pointed brackets to denote ordered pairs, e.g., $\langle x , y \rangle$, whereas intervals are given in the traditional way with brackets or parentheses, e.g., $(a,b), [c,d]$, etc.) The set-valued function $f: X \rightarrow 2^{Y}$ is said to be \emph{surjective} if, for every $y \in Y$, there exists $x \in X$ with $y \in f(x)$. If a u.s.c. function $f: X \rightarrow 2^{Y}$ has the property that $f(x)$ is connected for each $x \in X$, then $f$ is \emph{continuum-valued}, and we denote this by writing  $f: X \rightarrow C(Y).$ The \emph{identity on} $X$ (also called the \emph{trivial} function on $X$) is the u.s.c. function $\iota$ given by $\iota(x) = \{x\}$ for all $x \in X$; we may simply say the \emph{identity} (or, more simply, $\iota$) when the space $X$ is understood from context.

@@

Because it can be difficult to check if a large collection of u.s.c. functions is exact, a much more tractable case would be that of an inverse limit system with a single idempotent bonding function $f$. That is, suppose $X$ is a non-empty compact Hausdorff space and $f: X \rightarrow 2^{X}$ is an idempotent u.s.c. function. If $X_{\alpha} = X$ for each $\alpha \in \kappa$ and $f_{\alpha \beta} = f$ for each $\alpha < \beta \in \kappa$, then $\{X_{\alpha}, f_{\alpha \beta}, \kappa\} = \{X, f, \kappa\}$ is an \emph{inverse limit system with the single bonding function $f$}, and we denote the inverse limit of the system as $\varprojlim \{X, f, \kappa\}$. (Of course, since $f$ is idempotent, the collection $\textbf{f}$ of bonding functions is automatically exact.) We note that, although the only continuous idempotent surjective function from a compact Hausdorff space $X$ to itself is the identity map, there are many nontrivial examples of u.s.c. idempotent surjective functions on $X$; see Figure \ref{idempotentUsc} for examples.

@@

By applying results by Ingram and Mahavier \cite{ingram mahavier} and by Charatonik and Roe \cite{char roe}, Varagona stated theorems (2.1 and 2.2 in \cite{varagona}) that imply $\varprojlim \{[0,1], f, \kappa\}$ is a continuum when $f$ is u.s.c., idempotent, and continuum-valued (and $\kappa > 0$). On the other hand, Greenwood and Kennedy showed in \cite{greenwood kennedy} that there is a non-continuum-valued u.s.c. surjective idempotent bonding function $f$ (with a connected graph) for which $\varprojlim \{[0,1], f, \omega\}$ is disconnected.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{idempotent_usc_4.pdf}
\end{center}
\caption{Idempotent u.s.c. functions with highlighted points illustrating condition $\Gamma$}
\label{conditionGamma}
\end{figure}

@@

Note that while we have assumed that $f:[0,1]\to C([0,1])$ (which also guarantees that the inverse limit is a continuum), it is an open question whether $\Gamma$ must also hold for any nontrivial surjective idempotent u.s.c. $f:[0,1]\to 2^{[0,1]}$ (regardless of whether the inverse limit remains connected).

@@

Note that when $\kappa\geq\omega_1$, $\omega_1\in\kappa+1$ is a point of non-first-countability; therefore, any space containing $\kappa+1$ as a subspace is nonmetrizable.

@@

All compact metric spaces are embeddable in $\mathbb R^\omega=\Sigma\mathbb R^\omega$ and therefore Corson compact. In addition, every closed subspace of a Corson compact space is Corson compact. While $\Sigma\mathbb R^\kappa$ is not first-countable for $\kappa\geq\omega_1$, any subspace of $\Sigma\mathbb R^\kappa$ has the weaker property of being \textit{Fr\'echet-Urysohn}: each limit point of a set is converged to by a countable sequence in the set. However, $\omega_1+1$ is not Fr\'echet-Urysohn, as every countable sequence of ordinals less than $\omega_1$ is bounded above and therefore cannot converge to $\omega_1$ itself. Therefore, any space containing $\kappa+1$ for $\kappa\geq\omega_1$ cannot even be Corson compact.

@@

\bibitem{greenwood kennedy} Greenwood, Sina and Kennedy, Judy, {\it Connected generalized inverse limits}, Topology and its Applications,
159 (2012), no. 1, 57-68.
